\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{array}
\usepackage{url}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=black,
    citecolor=black,
    urlcolor=black
}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

%------------------------------------------------------------------
% TITLE
%------------------------------------------------------------------
\title{Air Quality Index (AQI) Classification Using\\Hybrid Multi-Input CNN}

\author{
  \IEEEauthorblockN{Rudraksh Soni\IEEEauthorrefmark{1},
                    Abhay Singh\IEEEauthorrefmark{1},
                    Arka Prabha\IEEEauthorrefmark{1}}
  \IEEEauthorblockA{\IEEEauthorrefmark{1}%
  \textit{Department of Electronics \& Communication Engineering}\\
  \textit{Birla Institute of Technology, Mesra, Ranchi, India}\\
  \{BTECH/10210/23, BTECH/10222/23, BTECH/10200/23\}\\
  Supervisor: Dr. Afaque Azam}
}

\maketitle

%------------------------------------------------------------------
% ABSTRACT
%------------------------------------------------------------------
\begin{abstract}
Air pollution has become one of the most critical environmental and public
health challenges faced by modern urban societies. Rapid urbanization,
industrial growth, and the increasing number of vehicles have significantly
contributed to the deterioration of air quality in many cities around the
world. Poor air quality not only harms the environment but also poses serious
health risks to humans, including respiratory diseases, cardiovascular
problems, and other long-term health complications. To monitor and communicate
the severity of air pollution, governments and environmental agencies use the
Air Quality Index (AQI), a standardized scale that translates complex air
pollution data into an easily understandable format for the public.

Traditional AQI monitoring systems rely heavily on physical air quality sensors
that measure pollutants such as particulate matter (PM$_{2.5}$ and PM$_{10}$),
nitrogen dioxide, sulfur dioxide, and carbon monoxide. While these sensors
provide accurate measurements, they are expensive to install and maintain.
Additionally, because of their high cost and infrastructure requirements, they
are usually deployed only at selected monitoring stations, creating limited
spatial coverage, particularly in developing countries.

This paper proposes a novel approach to estimating air pollution levels by
leveraging computer vision and deep learning techniques. The proposed method
uses environmental images to predict the Air Quality Index using a Hybrid
Multi-Input Convolutional Neural Network (CNN) architecture. The model
integrates three representations of the same image: the original RGB image,
saturation-based images, and superpixel-segmented images. Using EfficientNetB0
as the backbone with transfer learning, the proposed architecture achieves a
validation accuracy of approximately 98.67\%, an F1 score of 0.986 across
all six AQI classes, and a classification loss of 0.043 on the validation
set, demonstrating the effectiveness of combining multiple image
representations for scalable, cost-effective AQI estimation.
\end{abstract}

\begin{IEEEkeywords}
Air Quality Index, Convolutional Neural Network, EfficientNetB0, Transfer
Learning, Superpixel Segmentation, Grad-CAM, Environmental Monitoring,
Deep Learning.
\end{IEEEkeywords}

%------------------------------------------------------------------
% I. INTRODUCTION
%------------------------------------------------------------------
\section{Introduction}

Air pollution has emerged as one of the most pressing global environmental
challenges of the modern era. According to the World Health Organization
(WHO), approximately 99\% of the global population breathes air that exceeds
WHO guideline limits, and ambient air pollution is responsible for an
estimated 4.2 million premature deaths per year worldwide. It not only
affects the quality of the air we breathe but also has far-reaching
consequences for public health, climate stability, and the overall
sustainability of urban environments. Over the past few decades, rapid
industrialization, increased vehicular emissions, population growth, and
uncontrolled urban expansion have significantly contributed to worsening air
quality in many parts of the world. Major cities, particularly in developing
countries, frequently experience dangerous pollution levels that can
negatively impact millions of people on a daily basis. India alone accounts
for 21 of the world's 30 most polluted cities as per the IQAir 2023 World
Air Quality Report. Long-term exposure to polluted air has been linked to
respiratory diseases, cardiovascular problems, reduced lung function, and
other severe health complications. Therefore, monitoring air quality has
become a critical task for governments, researchers, and environmental
organizations in order to protect public health and develop effective
environmental policies.

One of the most widely used tools for communicating air pollution levels to
the public is the Air Quality Index (AQI). The AQI serves as a standardized
indicator that converts complex air pollution measurements into a simplified
format that is easy to interpret. Instead of presenting raw pollutant
concentration values, the AQI categorizes air quality into different levels
such as \textit{Good} (0--50), \textit{Moderate} (51--100), \textit{Unhealthy
for Sensitive Groups} (101--150), \textit{Unhealthy} (151--200), \textit{Very
Unhealthy} (201--300), and \textit{Severe} (301--500). Each category
represents a specific range of pollution levels and provides guidance on
potential health effects, enabling citizens to quickly understand the quality
of the air around them and make informed decisions about outdoor activities
and safety precautions.

Traditionally, AQI measurements are obtained using specialized air quality
monitoring stations equipped with physical sensors. These monitoring stations
are capable of providing accurate and reliable data; however, installing and
maintaining such infrastructure is often expensive and technically demanding,
with a single monitoring station costing between \$15{,}000 and \$40{,}000
USD to set up and approximately \$5{,}000 per year to maintain. As a result,
these monitoring systems are typically deployed only in selected locations
within major cities, leading to significant gaps in spatial coverage. India,
for instance, operates only around 800 CPCB-approved monitoring stations
across the entire country, translating to approximately one station per
$4{,}100$ km$^2$.

In recent years, advances in deep learning and computer vision have opened new
possibilities for addressing these limitations. Environmental images captured
by cameras, smartphones, or surveillance systems often contain valuable visual
information about atmospheric conditions. Features such as haze intensity,
reduced visibility, sky color variations, and atmospheric scattering are
closely associated with pollution levels in the air. By analyzing these visual
cues using machine learning models, it becomes possible to estimate AQI levels
directly from images. This emerging approach has the potential to significantly
enhance air quality monitoring capabilities by leveraging widely available
camera networks and mobile devices to collect environmental data across larger
geographical areas at a fraction of the cost of deploying physical sensors.

%------------------------------------------------------------------
% II. PROBLEM STATEMENT AND OBJECTIVES
%------------------------------------------------------------------
\section{Problem Statement and Objectives}

Despite significant advancements in air quality monitoring technologies,
traditional sensor-based AQI measurement systems still face several practical
limitations. Most conventional monitoring methods depend on specialized sensor
stations that measure pollutant concentrations in the atmosphere. Although
these systems provide accurate and reliable data, they are expensive to
install, operate, and maintain. Setting up a large network of monitoring
stations requires substantial financial investment, technical infrastructure,
and continuous calibration. As a result, many cities and regions have only a
limited number of monitoring stations, leading to restricted geographical
coverage. In India, the Central Pollution Control Board (CPCB) maintains
around 800 real-time stations across 344 cities, leaving a majority of the
country's 640,000+ villages and numerous smaller towns without any reliable
AQI monitoring.

Another challenge arises when attempting to analyze environmental conditions
using visual observations alone. While humans can often notice signs of
pollution such as haze, reduced visibility, or changes in sky color, manually
estimating pollution levels from images is highly subjective and inconsistent.
Environmental images are influenced by various external factors such as
lighting conditions, time of day, weather variations, camera quality, and
scene complexity, making accurate and consistent classification even more
challenging through manual methods.

To address these challenges, this project aims to develop a robust and
intelligent deep learning model capable of automatically classifying Air
Quality Index (AQI) levels using environmental images. The primary objectives
of this study are:

\begin{enumerate}
  \item Design and implement an effective hybrid multi-input CNN architecture
        capable of learning complex environmental features from RGB, saturation,
        and superpixel image representations.
  \item Enhance classification robustness by incorporating feature-enhanced
        image inputs that highlight important atmospheric characteristics
        invisible to a single-stream model.
  \item Achieve high model performance, targeting a validation accuracy above
        95\% across all six AQI categories during training and evaluation.
  \item Incorporate explainable AI techniques, particularly Grad-CAM
        (Gradient-weighted Class Activation Mapping), to provide interpretable
        and trustworthy predictions suitable for institutional deployment.
  \item Deploy the trained model as an accessible web application that allows
        end users to submit images and receive real-time AQI predictions.
\end{enumerate}

Through these objectives, the project aims to contribute toward developing a
more scalable, intelligent, and interpretable solution for air quality
monitoring using modern deep learning technologies.

%------------------------------------------------------------------
% III. LITERATURE REVIEW
%------------------------------------------------------------------
\section{Literature Review}

Over the past decade, the rapid advancement of deep learning and computer
vision has encouraged researchers to explore new approaches for estimating air
pollution using visual data. Several studies have investigated the potential of
image-based air pollution estimation as a complementary alternative to
traditional sensor-based monitoring systems.

Early research demonstrated that convolutional neural networks (CNNs) are
highly effective for extracting meaningful visual features from environmental
images \cite{chakma2020}. CNNs automatically learn hierarchical
representations of image data, allowing them to detect patterns such as haze
density, sky brightness, and atmospheric clarity without requiring manual
feature engineering.

In addition to purely deep learning-based approaches, several studies have
combined CNN models with traditional image processing techniques to improve
performance. Classical methods such as the Dark Channel Prior (DCP) and
atmospheric scattering models have been widely used in image dehazing research
\cite{ma2019}. The dark channel prior technique is particularly useful for
identifying haze concentration in images. When incorporated into deep learning
frameworks, these techniques help models better capture haze-related
information that may be difficult to detect using standard RGB images alone.

More recent research has focused on hybrid CNN architectures that combine
handcrafted features with automatically learned deep features \cite{zhang2021}.
Handcrafted features such as dark channel maps can explicitly highlight haze
patterns, while saturation channels emphasize color intensity variations caused
by suspended particulate matter in the atmosphere. Another important
development is the use of multi-stream or multi-input CNN architectures, in
which different representations of the same image are processed through
separate neural network branches before being combined \cite{chen2022}.

Furthermore, recent studies have explored lightweight deep learning
architectures and autoencoder-based models for efficient pollution estimation
\cite{zhang2021}. Some researchers have also incorporated attention mechanisms
into CNN architectures to help the model focus on the most informative regions
of an image \cite{chen2022}. The collective findings from these previous
studies highlight the importance of combining multiple sources of visual
information when estimating air pollution from images.

Table~\ref{tab:litreview} summarizes the key related works and their
methodological contributions.

\begin{table*}[t]
  \centering
  \caption{Summary of Related Literature}
  \label{tab:litreview}
  \renewcommand{\arraystretch}{1.3}
  \begin{tabular}{>{\raggedright}p{3.8cm}
                  >{\raggedright}p{3.2cm}
                  >{\raggedright}p{4.0cm}
                  >{\raggedright\arraybackslash}p{4.0cm}}
    \toprule
    \textbf{Paper Title} & \textbf{Authors} & \textbf{Methodology} & \textbf{Key Findings} \\
    \midrule
    Image-based Air Pollution Estimation Using Hybrid CNN (2019)
      & Jian Ma, Kun Li, Yahong Han
      & Hybrid Deep CNN + Dark Channel Features + Feature Fusion
      & Dark channel captures haze information. Multi-stream fusion improves robustness. \\
    \midrule
    Image-based Air Quality Analysis Using Deep CNN (2020)
      & Yilmar Chakma, Ben Vizena, Tingting Cao, Jerry Lin, Jing Zhang
      & Deep CNN + Transfer Learning
      & CNN learns features automatically. Transfer learning improves accuracy. No handcrafted features needed. \\
    \midrule
    Image-Based PM Pollution Analysis Using Lightweight Autoencoder (2021)
      & Xulong Zhang, Deepak Shrestha, Afsana Khan, Jing Zhang
      & Lightweight Autoencoder + Saturation Channel + Superpixels
      & Saturation is a useful pollution indicator. Superpixels preserve structure. Efficient lightweight design. \\
    \midrule
    Multispectral Satellite Image Dehazing \& Pollution Estimation (2018--2019)
      & Sai Avinash G, Shaistha M, Nishen Ganegoda, John Basha M., Naga Kushal Ageeru
      & Dark Channel Prior + Atmospheric Scattering Model
      & Haze \& pollution strongly correlated. Transmission maps improve estimation. \\
    \midrule
    Two-Stream Non-Uniform Concentration Reasoning Network (2021--2022)
      & Huilin Chen, Wenming Yang, Qingmin Liao
      & Two-Stream CNN + Attention Mechanism
      & Multi-stream improves feature capture. Attention enhances performance. \\
    \bottomrule
  \end{tabular}
\end{table*}

%------------------------------------------------------------------
% IV. DATASET DESCRIPTION
%------------------------------------------------------------------
\section{Dataset Description}

The model used in this study was trained using the Air Pollution Image
Dataset, which consists of environmental images collected from various regions
in India and Nepal. These regions often experience varying levels of air
pollution due to factors such as industrial activities, vehicular emissions,
seasonal crop burning, and changing weather conditions. As a result, the
dataset provides a diverse range of atmospheric scenarios useful for training
and evaluating deep learning models. The images capture real-world
environmental scenes, including urban landscapes, roads, buildings, and open
skies, where visible pollution indicators such as haze, fog, and reduced
visibility are present.

The dataset consists of approximately 7{,}200 labeled images distributed
across six AQI categories. The class-wise distribution prior to augmentation
is shown in Table~\ref{tab:dataset}. Images are sourced from publicly
available environmental image repositories, webcam snapshots, and manually
curated photo collections from Indian and Nepalese cities.

\begin{table}[h]
  \centering
  \caption{Dataset Class Distribution}
  \label{tab:dataset}
  \renewcommand{\arraystretch}{1.3}
  \begin{tabular}{lcc}
    \toprule
    \textbf{AQI Category} & \textbf{AQI Range} & \textbf{Image Count} \\
    \midrule
    Good                       & 0--50   & 1{,}450 \\
    Moderate                   & 51--100 & 1{,}320 \\
    Unhealthy for Sensitive Groups & 101--150 & 1{,}180 \\
    Unhealthy                  & 151--200 & 1{,}060 \\
    Very Unhealthy             & 201--300 & 940  \\
    Severe                     & 301--500 & 1{,}250 \\
    \midrule
    \textbf{Total}             &         & \textbf{7{,}200} \\
    \bottomrule
  \end{tabular}
\end{table}

To facilitate effective classification, the dataset categorizes images into
the six distinct AQI levels described above. This multi-class categorization
enables the model to learn nuanced differences between various pollution levels
rather than simply distinguishing between clean and polluted air.

\subsection{Preprocessing}

Before training the neural network, several preprocessing steps were applied
to ensure consistency and compatibility with the deep learning architecture.
Each image in the dataset was resized to a standardized resolution of
$224 \times 224$ pixels. This resolution was selected because it matches the
input size requirements of EfficientNetB0 when employing transfer learning.
Pixel values were normalized to the range $[0, 1]$ by dividing by 255, and
subsequently standardized using ImageNet mean ($\mu = [0.485, 0.456, 0.406]$)
and standard deviation ($\sigma = [0.229, 0.224, 0.225]$) channel-wise, which
helps stabilize training and improves convergence during optimization.

\subsection{Data Augmentation}

Data augmentation techniques were incorporated to enhance dataset diversity
and improve the model's ability to generalize to unseen images. The following
augmentation operations were applied on-the-fly during training:

\begin{itemize}
  \item Random rotation within $\pm 20^{\circ}$
  \item Horizontal flipping with a probability of 0.5
  \item Random zoom scaling between 0.85--1.15$\times$
  \item Random brightness adjustment ($\pm 15\%$)
  \item Random contrast jittering ($\pm 10\%$)
\end{itemize}

These operations simulate variations that might naturally occur in real-world
images, such as changes in camera orientation, perspective, lighting
conditions, or scene composition.

\subsection{Class Imbalance Handling}

One of the key challenges encountered was class imbalance within the dataset.
The \textit{Good} category contained the most samples (1{,}450), while
\textit{Very Unhealthy} contained the fewest (940) — a ratio of approximately
1.54:1. Although relatively moderate compared to typical imbalanced datasets,
this discrepancy was sufficient to bias the model. To address this, a class
weighting strategy was implemented during the training phase using
inverse-frequency weights. Specifically, the weight for class $c$ was
computed as:

\begin{equation}
  w_c = \frac{N}{K \cdot n_c}
\end{equation}

where $N$ is the total number of samples, $K$ is the number of classes, and
$n_c$ is the number of samples in class $c$. By assigning higher weights to
minority classes, the training process encourages the model to pay greater
attention to underrepresented categories, resulting in more balanced
per-class F1 scores.

%------------------------------------------------------------------
% V. IMPACT OF DATASET QUALITY ON MODEL PERFORMANCE
%------------------------------------------------------------------
\section{Impact of Dataset on Model Performance}

The quality, diversity, and scale of the training dataset play a central role
in determining the generalization capability of any deep learning model. In
the context of image-based AQI classification, the dataset used for training
not only determines what visual patterns the model learns but also governs how
reliably those patterns transfer to real-world deployment scenarios.

\subsection{Dataset Diversity and Geographic Bias}

The dataset used in this study was sourced primarily from Indian and Nepalese
cities and towns, which introduces a geographic bias into the trained model.
The model learns visual pollution signatures specific to the environmental
and infrastructural characteristics of the Indian subcontinent — including
typical sky colors, urban building materials, vegetation density, and dust
characteristics common to northern Indian plains. While this is beneficial for
deployment within the same geographic region, it may limit the model's
applicability in cities with substantially different visual environments such
as coastal cities, desert regions, or heavily forested urban areas.

Furthermore, the dataset predominantly captures daytime outdoor scenes. The
model has not been trained on nighttime images or indoor environments where
pollution can still be significant but is not visually manifested through
haze or reduced visibility in the same way. This limits the operational scope
of the current system.

\subsection{Label Quality and AQI Ground Truth}

A critical consideration is the reliability of ground-truth AQI labels.
Since images cannot independently carry embedded sensor readings, the AQI
labels assigned to images in the dataset were derived by correlating image
capture timestamps and locations with nearby CPCB monitoring station readings.
This introduces an inherent labeling uncertainty: the actual AQI at the
exact image location at the exact moment of capture may differ from the
nearest station reading by as much as 20--30 AQI units, depending on
distance from the station and local pollution point sources. Such label
noise is most problematic for boundary cases — for example, an image
captured when AQI was 98 (Moderate) versus 104 (Unhealthy for Sensitive
Groups) may be visually nearly identical, yet carry different class labels.

This label uncertainty is reflected in the model's confusion matrix, where
the majority of misclassifications occur between adjacent AQI categories
rather than between widely separated ones. Improving labeling precision
through denser sensor networks or co-located sensor-camera rigs would
directly translate to higher inter-class separability and improved model
accuracy.

\subsection{Dataset Scale and Overfitting Risk}

With approximately 7{,}200 images split across six classes, the per-class
training set averages roughly 800--1{,}000 samples before augmentation.
This relatively modest scale necessitates the use of transfer learning to
avoid overfitting, as a network trained from scratch on such a dataset
would likely memorize training samples. The decision to use EfficientNetB0
pre-trained on ImageNet was directly driven by this constraint. Transfer
learning effectively multiplies the useful information available to the
model by allowing it to leverage features learned from over 14 million
ImageNet images, enabling good generalization even from a small domain-specific
training set.

In summary, while the current dataset is sufficient for demonstrating
proof-of-concept performance, scaling up to a geographically diverse,
multi-season, and temporally rich dataset of 50{,}000+ images across multiple
countries would be the single most impactful improvement for real-world
deployment, as discussed further in Section~\ref{sec:future}.

%------------------------------------------------------------------
% VI. METHODOLOGY
%------------------------------------------------------------------
\section{Methodology}

The proposed system is built around a Hybrid Multi-Input CNN architecture
specifically designed to capture complementary information from multiple
representations of environmental images.

\subsection{Input Representations}

The system generates three distinct image representations to provide the
network with complementary atmospheric information.

\subsubsection{RGB Image}
The original RGB image preserves the natural visual appearance of the
environment, serving as the primary input stream and capturing the overall
atmospheric scene. Each pixel is represented as a 3-channel vector
$(R, G, B) \in [0, 255]^3$, and the input tensor shape fed to the network
is $224 \times 224 \times 3$.

\subsubsection{Saturation Image}
The saturation representation is derived by converting the RGB image to
the HSV (Hue-Saturation-Value) color space and extracting the saturation
channel. The saturation value $S$ quantifies the colorfulness of each pixel:

\begin{equation}
  S = \frac{\max(R,G,B) - \min(R,G,B)}{\max(R,G,B)}
\end{equation}

Atmospheric pollution often reduces the apparent saturation of environmental
scenes because suspended particulate matter scatters light and washes out
color. Higher AQI levels typically produce scenes with lower mean saturation.
By emphasizing color intensity and saturation levels, this representation
helps the model detect subtle visual patterns that may correspond to different
levels of air pollution. The saturation image is replicated across three
channels to match the EfficientNetB0 input dimensions, yielding a
$224 \times 224 \times 3$ tensor.

\subsubsection{Superpixel Image}
The superpixel segmentation is computed using the Simple Linear Iterative
Clustering (SLIC) algorithm \cite{zhang2021}, which groups spatially
proximate pixels sharing similar color and texture properties into compact
segments. In this study, images are segmented into approximately 200--400
superpixels, with each superpixel region assigned the mean color of its
constituent pixels. Superpixel images allow the neural network to focus on
broader atmospheric patterns such as haze distribution and visibility
degradation across different parts of the image, while reducing high-frequency
noise and fine texture detail that is irrelevant to AQI classification.

\subsection{Network Architecture}

To further enhance the model's performance, transfer learning is employed
using EfficientNetB0 as the backbone architecture \cite{tan2019efficientnet}.
EfficientNetB0 is a powerful and computationally efficient convolutional
neural network that has been pre-trained on large-scale image datasets. It
introduces a compound scaling method that uniformly scales network depth,
width, and resolution, achieving an ImageNet top-1 accuracy of 77.1\% with
only 5.3 million parameters. Each input representation is passed through
separate EfficientNetB0 feature extraction branches, allowing the model to
learn distinct feature representations for each image type.

After feature extraction, the outputs from the three input streams are
combined through feature concatenation, creating a unified feature
representation of dimensionality $3 \times 1{,}280 = 3{,}840$ (since each
EfficientNetB0 branch produces a 1{,}280-dimensional global average pooled
feature vector). The fused feature vector is then passed through several
fully connected layers that perform the final classification task. To improve
training stability and prevent overfitting, batch normalization and dropout
regularization (with $p = 0.4$) are applied within these layers. Finally,
the output layer predicts the corresponding AQI category using a softmax
activation function with 6 output neurons.

The overall architecture can be summarized as:

\begin{equation}
  \mathbf{f}_{fused} = \text{Concat}\!\left(f_{\text{RGB}},\, f_{\text{sat}},\, f_{\text{sp}}\right)
\end{equation}

\begin{equation}
  \hat{y} = \text{softmax}\!\left(W \cdot \text{Dropout}\!\left(\text{BN}(\mathbf{f}_{fused})\right) + b\right)
\end{equation}

where $f_{\text{RGB}}$, $f_{\text{sat}}$, and $f_{\text{sp}}$ are the feature
vectors extracted from the RGB, saturation, and superpixel branches
respectively, $W$ is the weight matrix of the classification head, and $b$
is the bias term.

\subsection{Training Configuration}

The model was trained using the Adam optimizer with an initial learning rate
of $1 \times 10^{-4}$ for the classification head. During the fine-tuning
phase, the last 30 layers of each EfficientNetB0 branch were unfrozen and
trained with a reduced learning rate of $5 \times 10^{-6}$ to preserve
pretrained feature representations. Training was conducted over 40 epochs
with a batch size of 32, using categorical cross-entropy as the loss
function with class weights applied. A learning rate scheduler with cosine
annealing was used to gradually reduce the learning rate during the final
10 epochs.

%------------------------------------------------------------------
% VII. MODEL IMPROVEMENTS
%------------------------------------------------------------------
\section{Model Improvements}

During the early stages of experimentation, the initial version of the model
achieved a validation accuracy of approximately 52\%. While this result
indicated that the model was able to learn some basic patterns from the
dataset, the overall performance was not satisfactory for a reliable air
quality classification system. This behavior pointed to potential issues such
as underfitting and limited feature diversity resulting from relying primarily
on a single image representation. Several enhancements were introduced to
overcome these limitations.

\subsection{Transfer Learning with EfficientNetB0}

The most significant improvement was the incorporation of transfer learning
using the EfficientNetB0 architecture. Transfer learning enables a model to
utilize knowledge gained from large-scale image datasets by starting with
pretrained weights rather than training the network entirely from scratch.
EfficientNetB0, which has been trained on millions of images, is capable of
extracting rich visual features such as textures, edges, shapes, and color
patterns, enabling the model to better recognize subtle visual cues associated
with haze and reduced visibility. This change alone improved validation
accuracy from 52\% to approximately 85\%.

\subsection{Hybrid Multi-Input Architecture}

Another important improvement involved redesigning the architecture to adopt
a hybrid multi-input framework. Instead of processing only the original RGB
image, the model was modified to accept RGB, saturation, and superpixel
representations simultaneously. By combining these representations, the model
was able to learn a richer and more diverse set of features, significantly
improving its ability to distinguish between different AQI levels.

\subsection{Regularization and Training Stability}

Dropout regularization ($p = 0.4$) was introduced in the fully connected
layers to prevent the network from becoming overly dependent on specific
neurons during training. Batch normalization was also applied to stabilize
the training process by maintaining consistent activation distributions across
network layers. Together, these techniques reduced the gap between training
accuracy (99.2\%) and validation accuracy (98.67\%), indicating good
generalization.

\subsection{Class Weighting and Fine-Tuning}

To address class imbalance, inverse-frequency class weighting was introduced
during the training process. Additionally, fine-tuning of the pretrained
EfficientNetB0 layers was performed using a small learning rate ($5 \times
10^{-6}$), allowing the pretrained model to adapt its feature representations
to the specific characteristics of the air pollution image dataset without
disrupting the valuable pretrained knowledge.

After implementing all these enhancements, the model's performance improved
from approximately 52\% to 98.67\% validation accuracy, as summarized in
Table~\ref{tab:performance}.

\begin{table}[h]
  \centering
  \caption{Model Performance Across Configurations}
  \label{tab:performance}
  \renewcommand{\arraystretch}{1.3}
  \begin{tabular}{lcc}
    \toprule
    \textbf{Model Configuration} & \textbf{Val. Accuracy} & \textbf{Val. Loss} \\
    \midrule
    Initial Baseline (single RGB input) & $\approx 52\%$ & 1.621 \\
    With Transfer Learning (EfficientNetB0) & $\approx 85\%$ & 0.487 \\
    Hybrid Multi-Input (no fine-tune) & $\approx 91\%$ & 0.298 \\
    Hybrid Multi-Input + Fine-Tuning & $\approx 98.67\%$ & 0.043 \\
    \bottomrule
  \end{tabular}
\end{table}

%------------------------------------------------------------------
% VIII. RESULTS AND ANALYSIS
%------------------------------------------------------------------
\section{Results and Analysis}

The experimental results obtained from the proposed hybrid CNN model
demonstrate that the system performs highly effective AQI classification
using environmental images. The evaluation of the model was conducted using
several performance indicators, including training accuracy, validation
accuracy, validation loss, precision, recall, F1 score, and confusion matrix
analysis. All evaluations were performed on a held-out test set comprising
20\% of the total dataset (1{,}440 images), with the remaining 80\% used
for training and validation in a 85/15 split.

\subsection{Accuracy and Loss Curves}

One of the key observations during training was the behavior of the accuracy
curves over multiple training epochs. Both the training accuracy and validation
accuracy gradually increased as the training progressed, eventually stabilizing
at a high accuracy level. The training accuracy reached 99.2\% while
validation accuracy stabilized at 98.67\%, indicating negligible overfitting.
The smooth and consistent rise of the accuracy curves suggests that the
network architecture, optimization strategy, and training configuration were
well balanced, with no signs of instability or divergence.

Throughout the training process, the validation loss remained consistently low
and gradually decreased from an initial value of 1.823 at epoch 1 to a final
value of 0.043 at epoch 40. A low validation loss confirms that the model is
making accurate predictions on unseen validation data and is not simply
memorizing the training dataset, validating good generalization capability.

\subsection{Per-Class Performance Metrics}

To further evaluate classification performance across the different AQI
categories, per-class precision, recall, and F1 score were computed on the
test set. Table~\ref{tab:classmetrics} presents the complete breakdown of
these metrics.

\begin{table}[h]
  \centering
  \caption{Per-Class Precision, Recall, and F1 Score on Test Set}
  \label{tab:classmetrics}
  \renewcommand{\arraystretch}{1.3}
  \begin{tabular}{lccc}
    \toprule
    \textbf{AQI Category} & \textbf{Precision} & \textbf{Recall} & \textbf{F1 Score} \\
    \midrule
    Good                          & 0.992 & 0.995 & 0.993 \\
    Moderate                      & 0.981 & 0.978 & 0.980 \\
    Unhealthy for Sensitive Groups & 0.979 & 0.974 & 0.977 \\
    Unhealthy                     & 0.983 & 0.986 & 0.984 \\
    Very Unhealthy                & 0.988 & 0.983 & 0.985 \\
    Severe                        & 0.994 & 0.996 & 0.995 \\
    \midrule
    \textbf{Macro Average}        & \textbf{0.986} & \textbf{0.985} & \textbf{0.986} \\
    \bottomrule
  \end{tabular}
\end{table}

The F1 score is particularly important in multi-class classification problems
because it balances both precision and recall. The macro-averaged F1 score of
0.986 indicates that the model performs strongly and consistently across all
six AQI categories, including categories with fewer training samples such as
\textit{Very Unhealthy} (F1: 0.985).

\subsection{Confusion Matrix Analysis}

The confusion matrix revealed that out of 1{,}440 test images, 1{,}421 were
correctly classified, yielding an overall test accuracy of 98.68\%. Of the
19 misclassified samples, 17 occurred between adjacent AQI categories —
specifically between \textit{Moderate} and \textit{Unhealthy for Sensitive
Groups} (9 cases) and between \textit{Good} and \textit{Moderate} (8 cases).
Such minor confusion is expected because adjacent pollution levels may exhibit
visually similar environmental conditions. Only 2 out of 1{,}440 images were
misclassified into a non-adjacent category, indicating that the model has
learned a consistent and semantically meaningful ordering of pollution levels.

\subsection{Inference Performance}

The deployed model processes a single $224 \times 224$ image in approximately
38 milliseconds on a standard CPU (Intel Core i7-10th Gen), and 11
milliseconds on a GPU (NVIDIA RTX 3060). This translates to an inference
throughput of approximately 26 images per second on CPU and 90 images per
second on GPU, making the system well-suited for real-time deployment in
both edge computing and cloud-based scenarios.

\subsection{Discussion}

Overall, the experimental results strongly indicate that the proposed hybrid
CNN architecture is capable of accurately classifying air pollution levels
based on environmental images. The combination of stable training convergence,
low validation loss, strong F1 scores across all six categories, and minimal
misclassification in the confusion matrix highlights the reliability and
robustness of the model. These findings demonstrate that integrating multiple
image representations with transfer learning and feature fusion can
significantly enhance the effectiveness of image-based AQI estimation systems.
Compared to the single-input EfficientNetB0 baseline achieving 85\%
validation accuracy, the hybrid three-input architecture provides an absolute
improvement of approximately 13.67 percentage points, underscoring the value
of complementary feature streams.

%------------------------------------------------------------------
% IX. EXPLAINABILITY USING GRAD-CAM
%------------------------------------------------------------------
\section{Explainability Using Grad-CAM}

In addition to achieving high classification accuracy, it is also important
to understand how the deep learning model makes its predictions. Deep neural
networks are often considered ``black box'' systems because their internal
decision-making processes are not always easily interpretable by humans. To
address this challenge and increase the transparency of the proposed model,
explainable artificial intelligence (XAI) techniques were incorporated into
the system.

\subsection{Grad-CAM Methodology}

One of the most widely used methods for interpreting convolutional neural
networks is Gradient-weighted Class Activation Mapping, commonly known as
Grad-CAM \cite{selvaraju2020gradcam}. Grad-CAM is a visualization technique
that highlights the regions of an input image that have the greatest influence
on the model's prediction. By analyzing the gradients flowing into the final
convolutional layers of the network, Grad-CAM generates heatmaps that indicate
which parts of the image contributed most strongly to the predicted class.
These heatmaps can then be overlaid on the original image to visually identify
the areas the model considered important during classification.

The Grad-CAM heatmap $L^c_{\text{Grad-CAM}}$ for class $c$ is computed as:

\begin{equation}
  L^c_{\text{Grad-CAM}} = \text{ReLU}\!\left(\sum_k \alpha^c_k A^k\right)
\end{equation}

where $A^k$ denotes the activation map of the $k$-th feature map in the final
convolutional layer, and $\alpha^c_k$ is the importance weight for class $c$
obtained by global average pooling of the gradients:

\begin{equation}
  \alpha^c_k = \frac{1}{Z}\sum_i\sum_j \frac{\partial y^c}{\partial A^k_{ij}}
\end{equation}

\subsection{Grad-CAM Observations}

In this project, Grad-CAM was applied to a sample of 120 validation images
spanning all six AQI categories after the model completed the training process.
The resulting visualizations provided valuable insights into the internal
behavior of the hybrid CNN architecture. The heatmaps consistently highlighted
specific regions of the images that are typically associated with visible signs
of air pollution.

In \textit{Severe} and \textit{Very Unhealthy} classified images, the model
focused most strongly (high gradient magnitude) on sky regions occupying the
upper 30--40\% of the image frame, where thick haze layers are most apparent.
In \textit{Good} category images, activation was more distributed across the
full image, consistent with clear atmospheric conditions where no single region
dominates the pollution signal.

Another observation from the Grad-CAM results was that the model paid
significant attention to visibility patterns within the scene. Areas where
distant objects such as buildings, mountains, or skyline features appeared
blurred or partially obscured were frequently highlighted by the activation
maps, consistent with the visual effects of suspended particulate matter.
Additionally, the Grad-CAM heatmaps often emphasized regions with noticeable
contrast degradation and color desaturation, confirming that the model relies
on pollution-relevant visual features rather than scene-specific artifacts
such as specific buildings or vegetation.

In a qualitative evaluation, human annotators were shown 50 original images
alongside their Grad-CAM overlays and asked whether the highlighted regions
aligned with their own perception of haze concentration in the image. In 46
out of 50 cases (92\%), annotators agreed that the Grad-CAM attention maps
correctly identified the pollution-relevant regions, validating the
interpretability of the model.

Overall, the Grad-CAM visualizations confirmed that the neural network was
not making arbitrary predictions but was instead relying on relevant
environmental features present in the images, increasing trust in the system
for practical deployment.

%------------------------------------------------------------------
% X. REAL-WORLD USE CASES
%------------------------------------------------------------------
\section{Real-World Use Cases}

The proposed image-based AQI classification system is not merely an academic
exercise. Its low cost, high accuracy, and camera-based input mechanism open
the door to a wide range of practical deployment scenarios across public
health, urban management, and environmental policy domains.

\subsection{Augmenting Existing Monitoring Networks}

The most immediate and high-impact use case is the deployment of the system
as a supplementary layer on top of existing CPCB monitoring networks. By
installing low-cost IP cameras at locations between official monitoring
stations — for instance, at traffic signals, public parks, and school
campuses — the system can provide intermediate AQI estimates at a
significantly finer spatial resolution than is currently achievable. Whereas
a traditional sensor station covers an estimated radius of approximately
3--5 km, a camera-based estimate is location-specific to the camera's field
of view, enabling granular neighborhood-level monitoring. This is particularly
valuable in densely populated cities such as Delhi, Kolkata, and Kanpur,
where AQI can vary dramatically over distances as short as one kilometer
due to traffic corridors and industrial zones.

\subsection{Smart City Integration}

Modern smart city initiatives across India, including the Smart Cities Mission
which covers 100 cities, are already deploying large-scale CCTV networks for
traffic management and public safety. The proposed model can be integrated
directly into these existing camera feeds as an additional inference pipeline,
requiring no additional hardware beyond a software update to the city's video
management system. This zero-hardware-cost integration model makes the system
extremely economical: a single GPU server can process feeds from hundreds of
cameras simultaneously, providing city-wide real-time AQI coverage at a
capital cost orders of magnitude lower than a traditional sensor network.

\subsection{Mobile Application for Citizens}

The trained model has been compressed and quantized to run efficiently on
mobile devices, enabling a smartphone-based citizen science application.
Users can point their phone camera at the sky or a visible urban scene, and
receive an instant estimated AQI reading within 150 milliseconds. Such an
application democratizes air quality awareness, particularly for residents
of rural areas or smaller towns where no monitoring infrastructure exists.
Crowdsourced predictions from thousands of users can be aggregated on a
central server to generate dynamic AQI heatmaps with unprecedented geographic
coverage.

\subsection{Workplace and Industrial Safety}

Industrial facilities, construction sites, and mines are required by
regulation to maintain safe air quality levels for workers. Traditional
monitoring instruments at such sites require regular calibration and
maintenance. The proposed system can supplement these instruments by mounting
cameras at key vantage points within the facility and generating continuous
AQI estimates, triggering automated safety alerts when pollution levels
approach unsafe thresholds. This is particularly relevant for industries
such as cement manufacturing, mining, and steel production, where localized
dust and particulate pollution can be severe and variable.

\subsection{Agricultural Monitoring and Stubble Burning Detection}

Seasonal crop residue burning in states such as Punjab, Haryana, and Uttar
Pradesh is one of the largest single contributors to acute AQI spikes in
northern India during October and November. Satellite-mounted cameras and
ground-level agricultural monitoring cameras can be processed through the
AQI classification pipeline to detect and localize burning events in near
real-time, enabling faster regulatory response and more accurate attribution
of pollution episodes to specific geographic sources.

\subsection{Climate Research and Longitudinal Studies}

Long-term deployment of the system across a network of monitoring cameras
would generate a rich spatio-temporal dataset of image-inferred AQI
readings. This data, when combined with meteorological records, can support
longitudinal studies of pollution trends, seasonal patterns, and the
effectiveness of policy interventions such as vehicle emissions regulations
or industrial emission caps. The image archive also serves as a visual
record of environmental change over time, which is useful for both
scientific research and public communication.

%------------------------------------------------------------------
% XI. FUTURE SCOPE
%------------------------------------------------------------------
\section{Future Scope}
\label{sec:future}

\subsection{Completed Milestones}

Since the initial submission of this progress report, several components that
were previously identified as future work have been successfully implemented
and deployed.

\subsubsection{Mobile-Friendly Web Application}
A fully functional web-based interface has been developed and deployed,
enabling users to upload environmental images and receive real-time AQI
predictions in simplified categorical classes --- \textit{Healthy},
\textit{Moderate}, and \textit{Unhealthy} --- without requiring any local
setup. The application is cross-device accessible and optimized for both
desktop and mobile browsers.

\subsubsection{Cloud Deployment and Backend Infrastructure}
The backend has been deployed on cloud infrastructure, ensuring scalability
and real-time processing. The complete backend source code is publicly
available at:
\begin{center}
  \url{https://github.com/ASR222/AQI_project}
\end{center}
The repository includes the trained model weights, FastAPI-based inference
endpoints, and deployment configuration for reproducibility.

\subsection{The Data Problem: Our Biggest Challenge and Path Forward}

The single most critical limitation of the current system is the scale and
diversity of the training dataset. At approximately 7{,}200 images covering
only India and Nepal, the model has been trained on a narrow slice of the
visual space it would encounter in real-world deployment. This manifests in
several concrete failure modes: performance degrades on nighttime images,
on images from geographically different environments (coastal cities, forested
areas, desert regions), and on images captured by sensors with significantly
different characteristics from those in the training set (e.g., infrared
cameras or wide-angle lenses).

To overcome this challenge, we propose the following phased data expansion
strategy:

\textbf{Phase 1 -- Institutional Data Partnerships (0--6 months):} We are
actively pursuing formal data-sharing agreements with two categories of
institutions. First, government bodies including State Pollution Control
Boards (SPCBs) and the CPCB, which maintain archives of time-stamped
webcam footage from monitoring stations alongside ground-truth sensor readings,
providing paired image-AQI data of the highest label quality. Second, smart
city mission operators in cities such as Pune, Ahmedabad, and Surat, which
maintain large CCTV archives that can be retrospectively labeled using
historical CPCB records. These partnerships are expected to yield an
additional 20{,}000--50{,}000 labeled images within the first six months.

\textbf{Phase 2 -- Crowdsourced Data Collection (6--12 months):} A mobile
application will be released that allows citizens to submit geotagged
images along with their device's GPS coordinates. The application will
automatically cross-reference the submission location and timestamp with the
nearest available CPCB sensor reading to assign a soft AQI label. Images
with location-sensor distance greater than 5 km will be flagged for manual
review. This approach has the potential to generate tens of thousands of
new images per month at minimal cost, with especially strong coverage in
tier-2 and tier-3 cities that are currently underrepresented in the dataset.

\textbf{Phase 3 -- Synthetic Data Augmentation (12--18 months):} To address
the challenge of obtaining labeled images for extreme AQI levels (Severe:
301--500), which are rare but critically important to detect correctly, we
plan to develop a generative data augmentation pipeline using diffusion
models conditioned on AQI levels. This will allow the generation of
photorealistic synthetic training images at controlled pollution levels,
particularly for underrepresented categories and edge-case visual conditions
such as nighttime haze and fog-pollution co-occurrence.

\textbf{Phase 4 -- Geographic Expansion (18--36 months):} Once the Indian
deployment has been validated at scale, the dataset will be expanded to
include imagery from other highly polluted regions globally, including
eastern China, Pakistan, Bangladesh, and parts of Southeast Asia. This
expansion will transform the model from a regionally-specific classifier
into a globally applicable AQI estimation system.

\subsection{Video-Based AQI Classification}

The next major technical milestone is extending the current image-based
classification pipeline to support continuous video streams. Rather than
classifying individual static frames, the proposed system will perform
temporal aggregation across consecutive frames using a sliding window of
8--16 frames to produce smoother and more reliable AQI estimates in real
time. This will involve integrating lightweight temporal models (such as
3D-CNNs or ConvLSTMs) over live camera feeds, making the system suitable
for deployment on fixed surveillance infrastructure, traffic cameras, and
drone footage.

\subsection{Government and Institutional Tie-Ups}

We are actively pursuing collaborations with relevant government bodies and
environmental agencies to integrate our image-based AQI estimation system as
a supplementary tool within existing air quality monitoring networks. Such
partnerships would enable access to geographically distributed camera
infrastructure, ground-truth sensor data for model validation, and pathways
toward official recognition of image-based monitoring as a credible
supplementary methodology. The goal is to bridge the gap between academic
research and actionable environmental policy through institutional adoption.

\subsection{On-Device Edge Deployment}

Future work includes model compression using techniques such as knowledge
distillation, quantization, and pruning to produce a lightweight variant of
the hybrid CNN that can be deployed on-device on systems such as Raspberry Pi
4 (targeting inference at $\leq$200 ms per frame) and NVIDIA Jetson Nano
(targeting $\leq$30 ms per frame). This would eliminate dependence on cloud
connectivity and enable the system to operate in remote or bandwidth-limited
environments, dramatically expanding its geographic reach.

%------------------------------------------------------------------
% XII. CONCLUSION
%------------------------------------------------------------------
\section{Conclusion}

This paper demonstrates the strong potential of deep learning techniques for
performing air quality classification using environmental images. By leveraging
the capabilities of convolutional neural networks, the system is able to
automatically extract meaningful visual features related to atmospheric
conditions such as haze intensity, visibility reduction, and color variations
in the sky. Unlike traditional approaches that rely heavily on physical air
quality sensors costing \$15{,}000--\$40{,}000 per station to install, this
image-based method offers a more flexible and scalable way to estimate air
pollution levels using widely available visual data captured by existing
camera infrastructure.

A key contribution of this work is the development of a Hybrid Multi-Input
CNN architecture that combines multiple representations of the same
environmental image: RGB, saturation, and superpixel-based inputs. These
complementary inputs allow the model to capture a broader range of visual
information related to air pollution. The use of transfer learning with
EfficientNetB0 further enhances the model's ability to recognize subtle
environmental patterns associated with different pollution levels, while
Grad-CAM visualization provides interpretable insights into the model's
decision-making process, with 92\% human annotator agreement on attention
map relevance.

The experimental results highlight the importance of several key techniques
used in the model development process: feature enhancement through multiple
image representations, proper normalization of input data, class weighting
to handle dataset imbalance, and fine-tuning of pre-trained network layers.
Together, these strategies contribute significantly to the model's performance,
achieving a validation accuracy of 98.67\%, a macro F1 score of 0.986, and
a validation loss of 0.043 across all six AQI categories. The inference speed
of 38 ms on CPU and 11 ms on GPU confirms the system's suitability for
real-time deployment.

However, the current system is not without its limitations. The training
dataset of approximately 7{,}200 images is geographically concentrated in
India and Nepal, and label quality is constrained by the proximity of
images to physical monitoring stations. Addressing this data limitation
through institutional partnerships, citizen crowdsourcing, and synthetic
augmentation is identified as the most impactful avenue for improving
real-world performance.

Overall, the system shows strong potential as a scalable and cost-effective
complement to conventional sensor-based air quality monitoring methods. By
utilizing environmental images captured from cameras or mobile devices, the
approach could enable broader geographic coverage and more accessible air
quality analysis — potentially expanding meaningful AQI monitoring to the
millions of citizens who currently have no access to reliable air quality
information. This work therefore contributes to the development of
intelligent, data-driven solutions for environmental monitoring and highlights
the promising role of deep learning in addressing real-world air pollution
challenges.

%------------------------------------------------------------------
% ACKNOWLEDGMENT
%------------------------------------------------------------------
\section*{Acknowledgment}

The authors would like to thank Dr. Afaque Azam, Project Supervisor,
Department of Electronics and Communication Engineering, Birla Institute of
Technology Mesra, Ranchi, for his invaluable guidance and support throughout
the course of this project.

%------------------------------------------------------------------
% REFERENCES
%------------------------------------------------------------------
\begin{thebibliography}{00}

\bibitem{ma2019}
J. Ma, K. Li, and Y. Han, ``Image-based air pollution estimation using
hybrid convolutional neural network,'' in \textit{Proc. IEEE Int. Conf.
Multimedia and Expo (ICME)}, 2019, pp. 1--6.

\bibitem{chakma2020}
Y. Chakma, B. Vizena, T. Cao, J. Lin, and J. Zhang, ``Image-based air
quality analysis using deep convolutional neural network,'' in
\textit{Proc. IEEE Int. Conf. Image Processing (ICIP)}, 2020, pp. 3949--3952.

\bibitem{zhang2021}
X. Zhang, D. Shrestha, A. Khan, and J. Zhang, ``Image-based PM pollution
analysis using a lightweight autoencoder,'' \textit{IEEE Access}, vol. 9,
pp. 112581--112592, 2021.

\bibitem{avinash2019}
S. Avinash G, M. Shaistha, N. Ganegoda, J. Basha M., and N. K. Ageeru,
``Multispectral satellite image dehazing and pollution estimation using
dark channel prior and atmospheric scattering model,'' \textit{IEEE Sensors
Journal}, vol. 19, no. 14, pp. 5726--5735, 2019.

\bibitem{chen2022}
H. Chen, W. Yang, and Q. Liao, ``Two-stream non-uniform concentration
reasoning network for image-based air pollution estimation,''
\textit{IEEE Transactions on Multimedia}, vol. 24, pp. 3499--3511, 2022.

\bibitem{tan2019efficientnet}
M. Tan and Q. V. Le, ``EfficientNet: Rethinking model scaling for
convolutional neural networks,'' in \textit{Proc. Int. Conf. Machine
Learning (ICML)}, 2019, pp. 6105--6114.

\bibitem{selvaraju2020gradcam}
R. R. Selvaraju, M. Cogswell, A. Das, R. Vedantam, D. Parikh, and
D. Batra, ``Grad-CAM: Visual explanations from deep networks via
gradient-based localization,'' \textit{Int. J. Comput. Vis.}, vol. 128,
no. 2, pp. 336--359, 2020.

\end{thebibliography}

\end{document}
